---
title: "R Notebook"
---

## Bonus 2! 
Ok so at the end of the last R notebook I said it was the end of the series. And I've been talking about 3 R notebooks throughout the previous notebooks. 

But then Sarah asked me to make sure all of here data worked in R as well. 

Luckily Sarah's data does open and can be processed using all the same techniques as you have already learnt in the previous R notebooks. There are just a few extra steps for a couple of the files. Let's take a look! 

First, let's load one of the 2D NetCDF files. Luckily these can be be loaded in exactly the same way as the NetCDF files in the other R notebooks. The only extra thing to notice is that the variable holding the missing values is not "_FillValue" as it was in the NetCDF files used in R Notebook 2 and 3. The missing values variable in the NetCDF file here is called "missing_values". I found this out by reading the output of print(ncfile) in the usual way as taught in the previous R notebooks. 

```{r}
# Load the ncdf4 library into R. (note, we installed this with install.packages() in Notebook 2)
library("ncdf4")

# Open a file connection to the NetCDF file
ncfile <- nc_open("E:\\Murray\\R_netcdf-master\\R_netcdf-master\\140625.p0055c.x1p5CO2outgas\\140625.p0055c.x1p5CO2outgas\\biogem\\fields_biogem_2d.nc")

# Print the NetCDFs header
print(ncfile)

# Extract variables from the NetCDF file
lat = ncvar_get(ncfile, 'lat')
lon = ncvar_get(ncfile,'lon')
sed_det = ncvar_get(ncfile, 'sed_det')
sed_CaCO3 = ncvar_get(ncfile, 'sed_CaCO3')
carb_conc_CO3 = ncvar_get(ncfile, 'carb_conc_CO3')
grid_topo = ncvar_get(ncfile, 'grid_topo')

# Extract attributes from the NetCDF file
sed_det_units = ncatt_get(ncfile, 'sed_det', "units")
sed_CaCO3_units = ncatt_get(ncfile, 'sed_CaCO3', "units")
carb_conc_CO3_units = ncatt_get(ncfile, 'carb_conc_CO3', "units")
grid_topo_units = ncatt_get(ncfile, "grid_topo", "units")

sed_det_name = ncatt_get(ncfile, "sed_det", "long_name")
sed_CaCO3_name = ncatt_get(ncfile, "sed_CaCO3", "long_name")
carb_conc_CO3_name = ncatt_get(ncfile, "carb_conc_CO3", "long_name")
grid_topo_name = ncatt_get(ncfile, "grid_topo", "long_name")

sed_det_missing_value = ncatt_get(ncfile, "sed_det", "missing_value")
sed_CaCO3_missing_value = ncatt_get(ncfile, "sed_CaCO3", "missing_value")
carb_conc_CO3_missing_value = ncatt_get(ncfile, "carb_conc_CO3", "missing_value")
grid_topo_missing_value = ncatt_get(ncfile, "grid_topo", "missing_value")

# Close the NetCDF file connection
nc_close(ncfile)

# Fill missing values with NA 
sed_det[sed_det==sed_det_missing_value$value] <- NA
sed_CaCO3[sed_CaCO3==sed_CaCO3_missing_value$value] <- NA
carb_conc_CO3[carb_conc_CO3==carb_conc_CO3_missing_value$value] <- NA
grid_topo[grid_topo==grid_topo_missing_value$value] <- NA

```

Let's plot a quick map of the values of the 4 variables we have extracted. Remember our variables are currently a cube of data with dimensions latitude x longitude x time. We can get a single slice of them in the usual way, using the [latitude,longitude,time] square bracket notation. remember leaving one of the spaces empty selects all the data, so [,,1] selects all of the latitude, all of the longitude and just the first time value, to give us a time slice. 

```{r}
image.plot(lon, lat, grid_topo, main=grid_topo_name$value)
image.plot(lon, lat, sed_det[,,1], main=sed_det_name$value)
image.plot(lon, lat, sed_CaCO3[,,1], main=sed_CaCO3_name$value)
image.plot(lon, lat, carb_conc_CO3[,,1], main=carb_conc_CO3_name$value)
```

Second of Sarah's files are a bunch of ASCII files. These files are delimited by a space, so can be read with the read.table function as in R Notebook 1. However the header in the file (telling us what each column corresponds to) has multiple words. This trips R over and we get an error message. This is because R looks at the first line to decide how many columns there are, and as R thinks every column has a space between it, then thinks there are too few columns. We can get around this using the "skip=1" argument, which tells R to skip the first row when reading in the data. We then use the "names()" function to manually change the names of the columns. 

```{r}
# Read in the data, skipping the header.
sed_det = read.table("E:\\Murray\\R_netcdf-master\\R_netcdf-master\\140625.p0055c.x1p5CO2outgas\\140625.p0055c.x1p5CO2outgas\\biogem\\biogem_series_sed_det.res", skip = 1)

# Set the names manually using the "names()" function
names(sed_det) = c("time", "ave_det_conc")

# Make a super basin plot to look at the data
plot(sed_det)

# Make a nicer plot
plot(sed_det, type="l", lwd=2, xlab="Time (yr)", ylab="Average detritacl concentration (wt%)")
grid(col='black')
```

Finally, let's look at one of Sarah's climate model outputs. This file has dimensions of latitude x longitude x depth x time. So it has one extra dimension. Luckily you handle this extra dimension in exactly the same way - with the square bracket notation. Let's see how:

```{r}
# Open a file connection to the NetCDF file
ncfile <- nc_open("E:\\Murray\\R_netcdf-master\\R_netcdf-master\\140625.p0055c.x1p5CO2outgas\\140625.p0055c.x1p5CO2outgas\\biogem\\fields_biogem_3d.nc")
# Print the NetCDFs header
print(ncfile)
# Extract variables from the NetCDF file
lat=ncvar_get(ncfile, 'lat')
lon=ncvar_get(ncfile,'lon')
misc_pH = ncvar_get(ncfile, 'misc_pH')

# Extract attributes from the NetCDF file
misc_pH_units = ncatt_get(ncfile, 'misc_pH', "units")
misc_pH_name=ncatt_get(ncfile, "misc_pH", "long_name")
misc_pH_missing_value=ncatt_get(ncfile, "misc_pH", "missing_value")

# Close the NetCDF file connection
nc_close(ncfile)
```

Let's look at the dimensions of the misc_pH variable
```{r}
dim(misc_pH)
```

Then lets extract a time and depth slice. This is just a slice of all latitudes, all longitudes, at 1 depth, at one time. 
```{r}
image.plot(lon, lat, misc_pH[,,1,1])
```

Finally, let's look at a method for changing the ticks around the plot. This is admitedly not very clear code, but it's an example of one of the few times that R is a bit of a pain!

```{r}
# First make extra space on the right hand side of the plot, for the colourbar
par(mar=c(5,4.5,4,7))
# Then plot the misc_pH matrix using image() not image.plot(). Use the axes=F argument to not plot any axes. 
image(lon, lat, misc_pH[,,1,1], axes=F, col=rainbow(256))
# Plot axis 1 (the bottom) with the defaults
axis(1)
# Plot axis 2 with ticks where we want them
axis(2, at=c(-60,-30,0,30,60))
# Use image.plot with the legend.only=T arguent to plot the colourbar only. 
image.plot(lon, lat, misc_pH[,,1,1], legend.only=T)
# Make the box around the figure. 
box()
```

